A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-01-31 analysis/paper/paper.Rmd
fix typo and add a citation

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns that can infer interactions and relationship between entities. Different from traditional static networks, statistical approaches to network analysis, such as exponential random graph models (ERGMs), provides a way to test hypotheses on dynamic process of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification for uncertainty, and robust model evaluation of network properties. We tested the hypothesis of social changes relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We explore social processes and network structures in archaeological data from from two time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to Exponential Random Graph Models (ERGMs) are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives a better convergence results. This enable us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. After fitting an ERMG with an MCMC algorithm, the Bayesian approach generates posterior probabilities incorporating our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by looking at the posterior mean and 95% credible intervals, or the odds ratios (conditional probability of observing an edge) that can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting an significant effect of a parameter is that its credible interval does not contain zero, or an odds ratio is greater/less than one, meaning a significant positive/negative effect. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data missing.

>>>>

---
title: "A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Burials provide valuable information to study social structures based on the assumption that burials and associated grave goods can represent social roles and relations in a society. To study social relationships, network analysis has been increasingly applied to archaeological data to explore network structures and patterns that can infer interactions and relationship between entities. Different from traditional static networks, statistical approaches to network analysis, such as exponential random graph models (ERGMs), provides a way to test hypotheses on dynamic process of network formation. However, the computational difficulty and sensitivity to uncertainties limits the application of ERGMs. In this paper, we introduce a Bayesian framework on ERGMs that enables an efficient computational process, effective quantification for uncertainty, and robust model evaluation of network properties. We tested the hypothesis of social changes relative to the arrival of Europeans by studying burial data from Kiwulan, an Iron Age site in northeastern Taiwan. The results indicate a transition from ties based on ritual objects to wealth objects, and a more centralized structure with social differentiation after European presence in the 17th century. Our case study demonstrates the effectiveness of Bayesian network analysis for archaeological data, and expands the use of burials in understanding the impacts of colonial presence on Indigenous groups in a pericolonial context.
keywords: |
  Bayesian modeling; network analysis; Monte-carlo simulation; Burials; European colonialism; social differentiation; Taiwan
highlights: |
  1. We present Bayesian inference on exponential random graph models (ERGMs) to model archaeological networks.
  2. We explore social processes and network structures in archaeological data from two time periods.
  3. Differences in network statistics imply change over time in social structures influenced by foreign presence. 
  4. Bayesian ERGMs are efficient and useful for evaluating and interpreting networks. 
---

@@

<!-- Bayesian inference for Exponential Random Graph Models --> 
Bayesian approaches to Exponential Random Graph Models (ERGMs) are effective tools for network modeling by incorporating prior information about the network configurations to better understand dependencies of network variables and improve computational issues in ERGMs [@Caimo2017; @Lehmann2020]. Prior information is derived from previous data or assumptions of our data. The advantage that Bayesian modeling has over traditional ERGMs is the application of Markov chain Monte Carlo (MCMC) simulation using the approximate exchange algorithm [@Caimo2011]. Bayesian ERGMs avoid doubly-intractable computations by directly sampling from the not normalized part of the posterior, which alleviates the computational problems and gives a better convergence results. This enable us to deal with complicated dependence patterns with ease, providing better estimations for complex social network models with heterogeneous data [@Caimo2017; @Snijders2006]. After fitting an ERMG with an MCMC algorithm, the Bayesian approach generates posterior probabilities incorporating our sample data and prior information that estimates the effect of each ERGM parameter in our models [@Caimo2015; @Nemmers2019]. This allows uncertainty quantification by looking at the posterior mean and 95% credible intervals, or the odds ratios (conditional probability of observing an edge) that can replace p-values for assessing a null hypothesis [@Caimo2017bayesian]. The typical criteria for interpreting an significant effect of a parameter is that its credible interval does not contain zero, or an odds ratio is greater/less than one, meaning a significant positive/negative effect [@Harris2013, pp. 51]. In addition, Bayesian approaches are also useful to deal with missing data, which is often a problem leading to misinterpretation of networks, especially for archaeological studies. @Koskinen2010 shows that the effect of missing data can be reduced with Bayesian modeling that can predict, on the average, 80% of the ties when a third of data missing.
